Data Specialization for a Staged Lambda Calculus

### 2015-07-03 paper/splitting.tex
work on the splitting and semantics sections

%\input{figures/splitting-overview}

@@

Splitting statically transforms terms in \lang\ into a pair of
monostaged terms ($f_1$ and $f_2(-)$) that correspond to the first-stage 
and second-stage computations in the original term. Since full
evaluation of world \bbtwo\ terms (both first-stage and second-stage
evaluation) only produces values in stage two, the idea of splitting
is to generate terms where evaluating $f_1$ in a first pass, and then
evaluating $f_2$ on the result of the first pass, yields the same
value as full evaluation of the original \lang\ program.

Similarly, when splitting \bbonem\ terms, first pass evaluation of
$f_1$ should produce the same value as first-stage evaluation of the
original \lang\ term.  However, complexity arises because
\bbonem\ terms may also contain computations to be evaluated in the
second stage. (For example, a \bbonem\ term of type $\fut \rmint$
results in an \rmint\ value in the second stage.) Thus we also want
the second pass evaluation of $f_2$ on the outputs of $f_1$ to yield
the same value as fully evaluating all second-stage computations in
the original \bbonem\ term. Finally, when splitting \bbonep\ terms,
which only produce values in the first stage, we only need to compare
the result of $f_1$ with that of the \lang\ term.

Handling the asymmetry between the results of evaluating
\bbonem\ terms (which may produce results at both the first and second
stage) and \bbtwo\ terms (which only produce second-stage results) is
fundamental to the structure of the splitting algorithm presented in
this section.

Our splitting algorithm is given by the judgements $\splittwo e A p l r$, $\splitone e A c {l.r}$, and $v \vsplito \mval i q$,
which apply to \bbtwo\ terms, \bbonem\ terms, and partial values, respectively.
Before discussing the meaning of these outputs, 
we state a few theorems regarding the totality and uniqueness of splitting:

\begin{theorem} \leavevmode
\label{thm:total} 
\begin{itemize} 
\item For term $e$, if $\typestwo e A$, then $\splittwo e A p l r$. 
\item For term $e$, if $\typesone e A$, then $\splitone e A p {l.r}$.
\item For partial value $v$, if $\typesone v A$, then $v \vsplito \mval i q$.
\end{itemize}
\end{theorem}

\begin{theorem} \leavevmode
\label{thm:unique}
\begin{itemize} 
\item If $\splittwo e A p l r$ and $\splittwo e A {p'} {l'} {r'}$, then $p = p'$, and $l.r = l'.r'$.
\item If $\splitone e A p {l.r}$ and $\splitone e A {p'} {l'.r'}$, then $p = p'$, and $l.r = l'.r'$.
\item If $v \vsplito \mval i q$ and $v \vsplito \mval {i'} {q'}$, then $i = i'$ and $q = q'$.
\end{itemize}
\end{theorem}

The proofs are by induction on the typing derivation
and simultaneous induction on the splitting derivations, respectively.

@@

\subsection{Term Splitting}

For any $\coltwo e A$
which reduces to the residual $\exv q$ (via repeated $\stepsym \bbtwo$),
splitting $e$ (via the new relation~$\splittwosym$)
yields a pair of monostaged terms $\pipeS p l r$, where $p$ and $l.r$
contain all first- and second-stage computations in $e$ respectively.

@@

We define the correctness of splitting 
by saying that $[b/l]r$, the \emph{applied resumer}, should be equivalent 
to the residual produced by direct evaluation, $q$.
More precisely:
\begin{theorem}
\label{thm:allCorrect}
If $\typestwo [\cdot] e A$, $e \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$, and $\splittwo e A p l r$,
then $\reduce p b$ and $[b/l]r \equiv q$.
\end{theorem}
In this theorem, ``$\reduce e v$'' indicates standard monostage reduction of the term $e$ to the value $v$,
and ``$\equiv$'' indicates a monostage equivalence, which we cover later.

We prove \ref{thm:allCorrect} by induction on the steps of evaluation.  
In the base case, where $e$ is already a residual of the form $\exv q$, we know
$\splittwo {\exv q} A {\exv{\tup{}}} \_ q$, and so by uniqueness of splitting, 
$p = b = \tup{}$ and $r = q$.
It directly follows that $\reduce {\exv {\tup {}}} {\tup {}}$ and $q \equiv q$.
In the recursive case, we have 
$\typestwo [\cdot] e A$, $e \stepsym \bbtwo e' \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$, and $\splittwo e A p l r$.
To help this case, we introduce to a new lemma, whose proof will concern the rest of this section:
\begin{lemma}
\label{lem:twoCorrect}
If $e \stepsym \bbtwo e'$, $\splittwo e A p  l r$, $\splittwo {e'} A {p'} {l'} {r'}$, and $\reduce {p'} {b'}$,
then $\reduce p b$ and $[b/l]r \equiv [b'/l']r'$.
\end{lemma}
By preservation and totality of splitting, we know $\typestwo [\cdot] {e'} A$ and $\splittwo {e'} A {p'} {l'} {r'}$.
The inductive hypothesis yields $\reduce {p'} {b'}$ and $[b'/l']r' \equiv q$.
From here, \ref{lem:twoCorrect} then gives $\reduce p b$ and $[b/l]r \equiv [b'/l']r'$,
and we derive $[b/l]r \equiv [b'/l']r' \equiv q$ directly. \qed

@@

For any $\colmix e A$, 
splitting $e$ (via the new relation~$\splitonesym$) again 
yields the pair of monostage terms $\pipeM c l r$. 
Evaluating $c$ in the first pass produces the pair
$\tup{i,b}$, where the first element $i$ represents the first stage part of the results of $e$,
and $b$ is the boundary value.
We can now define an analogue of \ref{lem:twoCorrect} for steps at \bbonem:
\begin{lemma}
\label{lem:oneCorrect}
If $e \stepsym \bbonem e'$, $\splitone e A p {l.r}$, $\splitone {e'} A {p'} {l'.r'}$, and $\reduce {c'} {\tup{i,b'}}$,
then $\reduce c {\tup {i,b}}$ and $[b/l]r \equiv [b'/l']r'$.
\end{lemma}
Like with the \bbtwo\ version, this requires that the applied resumer be equivalent both sides of the step.
But additionally, it also requires that first stages results be exactly identical.

\subsection{writing from here down...}
...

For example, in the simple case where $e$
contains entirely first-stage computation, such as:
\begin{lstlisting}
1` 3`gr{`1`literalone+literaltwo`3`}`
\end{lstlisting}
the first-stage result is the value 3.  The second element of the pair, $b$,
is the boundary value used as input to the resumer in second pass
evaluation as $[b/l]r \equiv q$, where $q$ is the same as
result produced by second-stage computations in $e$.  For example, in
the case where $\rtab \xi v$ contains entirely second-stage
computation, such as:
\begin{lstlisting}
2`[yhat |-> 2+3]`1` next{`2`yhat`1`}`
\end{lstlisting}
the second-stage result is the value 5. Since in general residual
tables and partial values describe a mixture of both first and second
stage computations, in the subsequent subsection we define an
operation, called {\em masking}, that separates residual tables and
partial values into their first- and second-stage results. We then
describe the algorithm for splitting both \bbonem\ terms
($\splitonesym$) and \bbtwo\ terms ($\splittwosym$).

%Since $c$ ensapulates computations that produce both $e$'s first-stage
%result and also the boundary value required by the resumer, we refer
%to it as a {\em combined term}.

%Note that since $\coltwo e A$ reduces to an entirely second-stage
%residual, its first-stage computations only exist to generate input
%for second-stage computations.  In contrast, since $\colmix e A$
%reduces to a multi-stage residual table and partial value, its first
%stage computations serve to both generate the value used as input for
%the second stage {\em and also} produce a first-stage value.
%Correspondingly, the precomputation $p$ resulting from splitting
%\bbtwo\ terms reduces only to a boundary value $b$ which is passed to
%the resumer; but the combined term $c$ resulting from splitting
%\bbonem\ terms reduces to a tuple containing both the first stage {\em
%  immediate result} $i$ and the boundary value $b$ (motivating the
%name {\em combined term}).

\subsection{Term Splitting at \bbonem}
\label{sec:split-one}

Given the definition of masking, we now describe how to split
\bbonem\ terms. The algorithm, given in \cref{fig:termSplit}, proceeds
recursively on the structure of~$e$. We derive a few cases here.

\paragraph {Unit.} 
As a simple first case, consider splitting the unit value, $\tup{}$
which masks to $\mval {\tup{}} {\tup{}}$.
%Units trivial contain unit split into more units: $\tup{} \vsplito \mval {\tup{}} {\tup{}}$.
Splitting must produce a $c$ and $l.r$ satisfying
\[
	\reduce c {\tup{\tup{},b}} \text{ and } [b/l]r \equiv \tup{}
\]
Our algorithm uses $c=\tup{\tup{},\tup{}}$ and $l.r = \_.\tup{}$.
Splitting $\pure{e}$ follows in similar fashion, except the first 
component of $c$ now becomes the result of evaluating the \bbonep\ term $e$.

\paragraph {Injections.}
Now consider splitting the non-terminal $\inl {e}$, where $\diaone
{e} \xi v$ and $\rtab \xi v \vsplito \mval i q$, and thus:
\[
	\splitone e A c {l.r} \text{ and } \reduce c {\tup{i,b}} \text{ and } [b/l]r \equiv q
\]
Since $\diaone {\inl e} \xi {\inl v}$ and $\rtab \xi {\inl v} \vsplito \mval {\inl i} q$,
splitting must produce a $c'$ and $l.r'$ satisfying
\[
	\reduce {c'} {\tup{\inl i,b'}} \text{ and } [b'/l]r' \equiv q
\]
Our algorithm generates a first pass $c'$ that produces the same
boundary value as splitting $e$:
\[
c' = \letin {\tup {x,y}} c {\tup{\inl x, y}}
\]
Splitting the injection also generates the same resumer as splitting
the subterm $e$ (that is, $l.r' = l.r$).

\paragraph{Tuple.} For an example of splitting terms with more than one argument, consider the tuple $\tup{e_1,e_2}$,
where $\diaone {e_k} {\xi_k} {v_k}$ and 
$\rtab {\xi_k} {v_k} \vsplito \mval {i_k} {q_k}$ for $k = 1,2$.  
Inductively:
\[
	\splitone {e_k} A {c_k} {l_k.r_k} \text{ and } \reduce {c_k} {\tup{i_k,b_k}} \text{ and } [b_k/l_k]r_k \equiv q_k
\]
Since $\diaone {\tup{e_i,e_2}} {\xi_1,\xi_2} {\tup{v_1,v_2}}$ 
and $\rtab {\xi_1,\xi_2} {\tup{v_1,v_2}}$ and $\tup{v_1,v_2} \vsplito \mval {\tup{i_1,i_2}} {\tup{q_1,q_2}}$,
splitting generates:
\[
	\reduce {c} {\tup{\tup{i_1,i_2},b}} \text{ and } [b/l]r \equiv \tup{q_1,q_2}
\]
The resulting boundary value is the tuple of the boundary values of
the subterms, $b = \tup{b_1,b_2}$.  The value is 
constructed using
\[
c = \letin{\tup{y_1,z_1}}{c_1}{\letin{\tup{y_2,z_2}}{c_2}{\tup{\tup{y_1, y_2},\tup{z_1, z_2}}}}
\]
and deconstructed in the resumer with a tuple pattern:
\[
l.r = \tup{l_1,l_2}.\tup{r_1,r_2}
\]
\paragraph {Case.}
Splitting is most complicated at $\caseof {e_1} {x_2.e_2} {x_3.e_3}$ terms because of their divergent control flow.
Here we consider only the case where
$\diaone {e_1} {\xi_1} {\inl {v_1}}$ and \\ $\rtab {\xi_1} {\inl {v_1}} \vsplito \mval {\inl {i_1}} {q_1}$.
(The \texttt{inr} case is analogous.)

\noindent
Inductively,
\[
	\splitone {e_1} A {c_1} {l_1.r_1} \text{ and } \reduce {c_1} {\tup{\inl {i_1},b_1}} \text{ and } [b_1/l_1]r_1 \equiv q_1
\]
Additionally, the first branch evaluates $\diaone {[v_1/x_2]e_2} {\xi_2} {v_2}$
and $\rtab {\xi_1,\xi_2} {v_2} \vsplito \mval {i_2} {q_2}$.
Therefore:
\begin{gather*}
	\splitone {e_2} A {c_2} {l_2.r_2} \text{ and } \reduce {[i_1/x_2]c_2}
  {\tup{i_2,b_2}} \\
\text{ and } [q_1/x_2][b_2/l_2]r_2 \equiv q_2
\end{gather*}
(Because $e_2$ is open on $x_2$, $\pipeM{c_2}{l_2}{r_2}$ is also open on $x_2$.
When we substitute some $v_1$ for $x_2$ in $e_2$, we must substitute the masks
of $v_1$ into $c_2$ and $l_2.r_2$.)

The entire case evaluates as
$\diaone {\caseof {e_1} {x_2.e_2} {x_3.e_3}} {\xi_1,\xi_2} {v_2}$, so splitting should produce 
some $c'$ and
$l'.r'$ such that 
\[
  \reduce {c'} {\tup{i_2,b'}} \text{ and }
  [b'/l']r' \equiv q_2
\]
We choose
\[
c' =
{\left(
\talllet{\tup{y_1,z_1}}{c_1}{
\tallcase{y_1}
{x_2.\letin{\tup{y_2,z_2}}{c_2}{\tup{y_2,\tup{z_1,\inl{z_2}}}}}
{x_3.\cdots\ttrpar}
}\right)}
\]
\[
l'.r' =
\tup{l_1,l_b}.\letin{z}{r_1}{\caseof{l_b}{l_2.[z/x_2]{r_2}}{l_3.\cdots}}
\]
The term $c'$ reduces to
$\tup{i_2,\tup{b_1,\inl{b_2}}}$, so 
$b' = \tup{b_1,\inl{b_2}}$. Therefore
$[b'/l']r'$ reduces to
$[[b_1/l_1]r_1/x_2][b_2/l_2]r_2$, exactly the 
$[q_1/x_2][b_2/l_2]r_2$ required.

More intuitively, $c'$ cases on $i_1$ to run the first branch of first-stage
code. The boundary value for that branch, $b_2$, is packaged as $\tup {b_1, \inl
{b_2}}$. The resumer of the predicate uses $b_1$, and $\inl{b_2}$ signals to run
the first branch's resumer. The second branches of $c'$ and $l'.r'$ are
similarly constrained by the $\texttt{inr}$ case.

%Recovering first stage divergence is easy, since for $\splitone {e_1} ? {c_1} {l_1.r_1}$, 
%$\reduce {e_1} {\inl {v_1}}$ implies $\reduce {c_1} {\tup{\inl {i_1},b_1}}$.

\paragraph {Function.} 
Function introduction has a $\tup{}$ boundary value, since functions
are already fully reduced in our semantics.  However, since the body
of a function may itself be multistage, splitting must descend into
it.  The first pass generated by splitting produces an $i$ that is a new
function formed from the first-stage part of the original body.  The
resumer is a new function formed out of the second-stage part of the
original body.  It is the responsibility of the application site to
save the precomputation of the function body and pass it to the
resumer version of the function.

\subsection{Term Splitting at \bbtwo}

Because world \bbtwo\ terms in \lang\ reduce to monostage residuals
(as opposed to partial values), term splitting at world
\bbtwo\ (specified by the $\splittwosym$ relation in
\cref{fig:termSplit}) is simpler than that of \bbonem.

Consider any $n$-ary \bbtwo-term $e = \scriptCapp {e_1 \ttsemi \ldots
  \ttsemi e_n}$ (except \texttt{prev}, which is discussed shortly).
Note that for any $\diatwo e q$, the residual $q$ must have the form
$\scriptCapp {q_1 \ttsemi \ldots \ttsemi q_n}$, even if $\mathcal{C}$ is
nullary or the $q_i$ have binders.
For each $k$, splitting the subterm $e_k$ yields $p_k$ and $l_k.r_k$ where
\[
	\reduce {p_k} {b_k} \text{ and } [{b_k}/{l_k}]{r_k} \equiv {q_k}
\]
Correspondingly, splitting $\scriptCapp {e_1 \ttsemi \ldots \ttsemi e_n}$ generates $p$ and $l.r$ where
\[
	\reduce p b \text{ and } [b/l]r \equiv \scriptCapp {q_1 \ttsemi \ldots \ttsemi q_n}
\]
Our algorithm accomplishes this by using $p = \tup {p_1, \ldots, p_n}$ and
$l.r = \tup {l_1, \ldots, l_n}.\scriptCapp {r_1 \ttsemi \ldots \ttsemi r_n}$.

%Again, this holds even for nullary terms and under binders.
In particular, $\caseof{e_1}{x_2.e_2}{x_3.e_3}$ splits into:
\[
\pipeS {\tup{p_1,p_2,p_3}} {\tup{l_1,l_2,l_3}} {\caseof{r_1}{x_2.r_2}{x_3.r_3}}
\]
Note how the precomputations of $e_2$ and $e_3$ are pulled out from
underneath the binder into a tuple that is always executed in the
first stage. This occurs regardless of what branch is eventually
executed in the second stage.

\paragraph{Staging constructs.}
Since \texttt{next} and \texttt{prev} do nothing more than convert between 
second-stage terms at \bbtwo\ and encapsulated second-stage terms at \bbonem,
their splitting rules are relatively simple.
Given $\diatwo e q$, splitting yields $p$ and $l.r$ where
\[
	\reduce p b \text{ and } [b/l]r \equiv q
\]
Then for $\diaone {\next {e}} {{\hat y} \mapsto q} {\next {\hat y}}$
and\\ $\rtab {{\hat y} \mapsto q} {\next {\hat y}} \vsplito \mval {\tup{}} {\letin {\hat y} q {\hat y}}$, splitting must produce $c'$ and $l'.r'$ where
\[
	\reduce {c'} {\tup {\tup{},b'}} \text{ and } [b'/l']r' \equiv \letin {\hat y} q {\hat y}
\]
Our algorithm chooses $c' = \tup{\tup{},p}$ and $l'.r' = l.r$.
Applying this process in reverse yields a splitting rule for $\prev e$ which applies \texttt{pi1} to the combined term of $e$.

>>>>

\input{figures/splitting-overview}

@@

% Splitting statically transforms terms in \lang\ into a pair of
% monostaged terms ($f_1$ and $f_2(-)$) that correspond to the first-stage 
% and second-stage computations in the original term.

% Recall that full evaluation (both the first and second pass) of a term $\coltwo e A$ 
% only produces its ``final result'' at the second stage.
% In contrast, \bbonem\ terms may have final results at both stages. 
% For example, consider the term:
% \begin{lstlisting}
% 1`(`3`gr{`1`4+5`3`}`1`, next{`2`6+7`1`})`
% \end{lstlisting}

% Thus, the idea of splitting for \bbtwo\ terms
% is to generate terms where evaluating $f_1$ in a first pass, and then
% evaluating $f_2$ on the result of the first pass, yields the same
% value as full evaluation of the original \lang\ program.

% first pass evaluation of
% $f_1$ should produce the same value as first-stage evaluation of the
% original \lang\ term.  However, complexity arises because
% \bbonem\ terms Thus we also want
% the second pass evaluation of $f_2$ on the outputs of $f_1$ to yield
% the same value as fully evaluating all second-stage computations in
% the original \bbonem\ term. Finally, when splitting \bbonep\ terms,
% which only produce values in the first stage, we only need to compare
% the result of $f_1$ with that of the \lang\ term.

% Handling the asymmetry between the results of evaluating
% \bbonem\ terms (which may produce results at both the first and second
% stage) and \bbtwo\ terms (which only produce second-stage results) is
% fundamental to the structure of the splitting algorithm presented in
% this section.

Our splitting algorithm is given by the judgements $\splittwo e A p l r$, $\splitone e A c {l.r}$, and $v \vsplito \mval i q$,
which apply to \bbtwo\ terms, \bbonem\ terms, and partial values, respectively.
Before discussing the meaning of these outputs, 
we note that splitting is defined for all well-typed inputs (\ref{thm:total} in \ref{fig:splittingSummary}), and it produces unique results (\ref{thm:unique}).
That is, each splitting relation is a total function.
We prove these theorems by straightforward induction on the typing derivation
and simultaneous induction on the splitting derivations, respectively.

@@

\subsection{World \bbtwo\ Term Splitting}

For any $\coltwo e A$
which reduces to the residual $\exv q$ (via repeated $\stepsym \bbtwo$),
splitting $e$ (via $\splittwosym$)
yields a pair of monostaged terms $\pipeS p l r$, where $p$ and $l.r$
contain all first- and second-stage computations in $e$ respectively.

@@

We define the correctness of splitting 
by saying that $[b/l]r$, the \emph{applied resumer}, should be equivalent 
to the residual produced by direct evaluation, $q$.
We state this more precisely as \ref{thm:allCorrect},
where ``$\reduce e v$'' indicates standard monostage reduction of the term $e$ to the value $v$,
and ``$\equiv$'' indicates a monostage equivalence, which we cover later.

We prove \ref{thm:allCorrect} by induction on the steps of evaluation.  
In the base case, where $e$ is already a residual of the form $\exv q$, we know
$\splittwo {\exv q} A {\exv{\tup{}}} \_ q$, and so by uniqueness of splitting, 
$p = \tup{}$ and $r = q$.
From here, we can directly derive $\reduce {\exv {\tup {}}} {\tup {}}$ and $q \equiv q$.

In the recursive case, we have 
$\typestwo [\cdot] e A$, $e \stepsym \bbtwo e' \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$, and $\splittwo e A p l r$.
To help this case, we introduce \ref{lem:twoCorrect}, whose proof will concern the rest of this section.
By preservation and totality of splitting, we know $\typestwo [\cdot] {e'} A$ and $\splittwo {e'} A {p'} {l'} {r'}$.
The inductive hypothesis yields $\reduce {p'} {b'}$ and $[b'/l']r' \equiv q$.
From here, \ref{lem:twoCorrect} then gives $\reduce p b$ and $[b/l]r \equiv [b'/l']r'$,
and we derive $[b/l]r \equiv [b'/l']r' \equiv q$ directly. \qed

@@

\subsection{World \bbonem\ Term Splitting}

\TODO why the added complexity for \bbonem\ terms?

For any $\colmix e A$, 
splitting $e$ (via $\splitonesym$) again 
yields the pair of monostage terms $\pipeM c l r$. 
Evaluating $c$ in the first pass produces the pair
$\tup{i,b}$, where the first element $i$ represents the first stage part of the results of $e$,
and $b$ is the boundary value.
Since execution of the first pass both serves to generate input for
the second pass, we say $c$ is a {\em combined term} that produces a
\emph{immediate result} ($i$) and a boundary value ($b$) for the resumer ($l.r$).

The correctness of \bbonem\ term splitting with respect to $\stepsym \bbonem$ is given in \ref{lem:oneCorrect}.
Like with the \bbtwo\ version, this requires that the applied resumer be equivalent both sides of the step.
But additionally, it also requires that first stages results be exactly identical.

\TODO talk about lifting step correctness

\TODO talk about ground case

\subsection{Example cases.}

\TODO do some cases of the proof, and use the opportunity to discuss why splitting does what it does

% For example, in the simple case where $e$
% contains entirely first-stage computation, such as:
% \begin{lstlisting}
% 1` 3`gr{`1`literalone+literaltwo`3`}`
% \end{lstlisting}
% the first-stage result is the value 3.  The second element of the pair, $b$,
% is the boundary value used as input to the resumer in second pass
% evaluation as $[b/l]r \equiv q$, where $q$ is the same as
% result produced by second-stage computations in $e$.  For example, in
% the case where $\rtab \xi v$ contains entirely second-stage
% computation, such as:
% \begin{lstlisting}
% 2`[yhat |-> 2+3]`1` next{`2`yhat`1`}`
% \end{lstlisting}
% the second-stage result is the value 5. Since in general residual
% tables and partial values describe a mixture of both first and second
% stage computations, in the subsequent subsection we define an
% operation, called {\em masking}, that separates residual tables and
% partial values into their first- and second-stage results. We then
% describe the algorithm for splitting both \bbonem\ terms
% ($\splitonesym$) and \bbtwo\ terms ($\splittwosym$).

% %Since $c$ ensapulates computations that produce both $e$'s first-stage
% %result and also the boundary value required by the resumer, we refer
% %to it as a {\em combined term}.

% %Note that since $\coltwo e A$ reduces to an entirely second-stage
% %residual, its first-stage computations only exist to generate input
% %for second-stage computations.  In contrast, since $\colmix e A$
% %reduces to a multi-stage residual table and partial value, its first
% %stage computations serve to both generate the value used as input for
% %the second stage {\em and also} produce a first-stage value.
% %Correspondingly, the precomputation $p$ resulting from splitting
% %\bbtwo\ terms reduces only to a boundary value $b$ which is passed to
% %the resumer; but the combined term $c$ resulting from splitting
% %\bbonem\ terms reduces to a tuple containing both the first stage {\em
% %  immediate result} $i$ and the boundary value $b$ (motivating the
% %name {\em combined term}).

% \subsection{Term Splitting at \bbonem}
% \label{sec:split-one}

% Given the definition of masking, we now describe how to split
% \bbonem\ terms. The algorithm, given in \cref{fig:termSplit}, proceeds
% recursively on the structure of~$e$. We derive a few cases here.

% \paragraph {Unit.} 
% As a simple first case, consider splitting the unit value, $\tup{}$
% which masks to $\mval {\tup{}} {\tup{}}$.
% %Units trivial contain unit split into more units: $\tup{} \vsplito \mval {\tup{}} {\tup{}}$.
% Splitting must produce a $c$ and $l.r$ satisfying
% \[
% 	\reduce c {\tup{\tup{},b}} \text{ and } [b/l]r \equiv \tup{}
% \]
% Our algorithm uses $c=\tup{\tup{},\tup{}}$ and $l.r = \_.\tup{}$.
% Splitting $\pure{e}$ follows in similar fashion, except the first 
% component of $c$ now becomes the result of evaluating the \bbonep\ term $e$.

% \paragraph {Injections.}
% Now consider splitting the non-terminal $\inl {e}$, where $\diaone
% {e} \xi v$ and $\rtab \xi v \vsplito \mval i q$, and thus:
% \[
% 	\splitone e A c {l.r} \text{ and } \reduce c {\tup{i,b}} \text{ and } [b/l]r \equiv q
% \]
% Since $\diaone {\inl e} \xi {\inl v}$ and $\rtab \xi {\inl v} \vsplito \mval {\inl i} q$,
% splitting must produce a $c'$ and $l.r'$ satisfying
% \[
% 	\reduce {c'} {\tup{\inl i,b'}} \text{ and } [b'/l]r' \equiv q
% \]
% Our algorithm generates a first pass $c'$ that produces the same
% boundary value as splitting $e$:
% \[
% c' = \letin {\tup {x,y}} c {\tup{\inl x, y}}
% \]
% Splitting the injection also generates the same resumer as splitting
% the subterm $e$ (that is, $l.r' = l.r$).

% \paragraph{Tuple.} For an example of splitting terms with more than one argument, consider the tuple $\tup{e_1,e_2}$,
% where $\diaone {e_k} {\xi_k} {v_k}$ and 
% $\rtab {\xi_k} {v_k} \vsplito \mval {i_k} {q_k}$ for $k = 1,2$.  
% Inductively:
% \[
% 	\splitone {e_k} A {c_k} {l_k.r_k} \text{ and } \reduce {c_k} {\tup{i_k,b_k}} \text{ and } [b_k/l_k]r_k \equiv q_k
% \]
% Since $\diaone {\tup{e_i,e_2}} {\xi_1,\xi_2} {\tup{v_1,v_2}}$ 
% and $\rtab {\xi_1,\xi_2} {\tup{v_1,v_2}}$ and $\tup{v_1,v_2} \vsplito \mval {\tup{i_1,i_2}} {\tup{q_1,q_2}}$,
% splitting generates:
% \[
% 	\reduce {c} {\tup{\tup{i_1,i_2},b}} \text{ and } [b/l]r \equiv \tup{q_1,q_2}
% \]
% The resulting boundary value is the tuple of the boundary values of
% the subterms, $b = \tup{b_1,b_2}$.  The value is 
% constructed using
% \[
% c = \letin{\tup{y_1,z_1}}{c_1}{\letin{\tup{y_2,z_2}}{c_2}{\tup{\tup{y_1, y_2},\tup{z_1, z_2}}}}
% \]
% and deconstructed in the resumer with a tuple pattern:
% \[
% l.r = \tup{l_1,l_2}.\tup{r_1,r_2}
% \]
% \paragraph {Case.}
% Splitting is most complicated at $\caseof {e_1} {x_2.e_2} {x_3.e_3}$ terms because of their divergent control flow.
% Here we consider only the case where
% $\diaone {e_1} {\xi_1} {\inl {v_1}}$ and \\ $\rtab {\xi_1} {\inl {v_1}} \vsplito \mval {\inl {i_1}} {q_1}$.
% (The \texttt{inr} case is analogous.)

% \noindent
% Inductively,
% \[
% 	\splitone {e_1} A {c_1} {l_1.r_1} \text{ and } \reduce {c_1} {\tup{\inl {i_1},b_1}} \text{ and } [b_1/l_1]r_1 \equiv q_1
% \]
% Additionally, the first branch evaluates $\diaone {[v_1/x_2]e_2} {\xi_2} {v_2}$
% and $\rtab {\xi_1,\xi_2} {v_2} \vsplito \mval {i_2} {q_2}$.
% Therefore:
% \begin{gather*}
% 	\splitone {e_2} A {c_2} {l_2.r_2} \text{ and } \reduce {[i_1/x_2]c_2}
%   {\tup{i_2,b_2}} \\
% \text{ and } [q_1/x_2][b_2/l_2]r_2 \equiv q_2
% \end{gather*}
% (Because $e_2$ is open on $x_2$, $\pipeM{c_2}{l_2}{r_2}$ is also open on $x_2$.
% When we substitute some $v_1$ for $x_2$ in $e_2$, we must substitute the masks
% of $v_1$ into $c_2$ and $l_2.r_2$.)

% The entire case evaluates as
% $\diaone {\caseof {e_1} {x_2.e_2} {x_3.e_3}} {\xi_1,\xi_2} {v_2}$, so splitting should produce 
% some $c'$ and
% $l'.r'$ such that 
% \[
%   \reduce {c'} {\tup{i_2,b'}} \text{ and }
%   [b'/l']r' \equiv q_2
% \]
% We choose
% \[
% c' =
% {\left(
% \talllet{\tup{y_1,z_1}}{c_1}{
% \tallcase{y_1}
% {x_2.\letin{\tup{y_2,z_2}}{c_2}{\tup{y_2,\tup{z_1,\inl{z_2}}}}}
% {x_3.\cdots\ttrpar}
% }\right)}
% \]
% \[
% l'.r' =
% \tup{l_1,l_b}.\letin{z}{r_1}{\caseof{l_b}{l_2.[z/x_2]{r_2}}{l_3.\cdots}}
% \]
% The term $c'$ reduces to
% $\tup{i_2,\tup{b_1,\inl{b_2}}}$, so 
% $b' = \tup{b_1,\inl{b_2}}$. Therefore
% $[b'/l']r'$ reduces to
% $[[b_1/l_1]r_1/x_2][b_2/l_2]r_2$, exactly the 
% $[q_1/x_2][b_2/l_2]r_2$ required.

% More intuitively, $c'$ cases on $i_1$ to run the first branch of first-stage
% code. The boundary value for that branch, $b_2$, is packaged as $\tup {b_1, \inl
% {b_2}}$. The resumer of the predicate uses $b_1$, and $\inl{b_2}$ signals to run
% the first branch's resumer. The second branches of $c'$ and $l'.r'$ are
% similarly constrained by the $\texttt{inr}$ case.

% %Recovering first stage divergence is easy, since for $\splitone {e_1} ? {c_1} {l_1.r_1}$, 
% %$\reduce {e_1} {\inl {v_1}}$ implies $\reduce {c_1} {\tup{\inl {i_1},b_1}}$.

% \paragraph {Function.} 
% Function introduction has a $\tup{}$ boundary value, since functions
% are already fully reduced in our semantics.  However, since the body
% of a function may itself be multistage, splitting must descend into
% it.  The first pass generated by splitting produces an $i$ that is a new
% function formed from the first-stage part of the original body.  The
% resumer is a new function formed out of the second-stage part of the
% original body.  It is the responsibility of the application site to
% save the precomputation of the function body and pass it to the
% resumer version of the function.

% \subsection{Term Splitting at \bbtwo}

% Because world \bbtwo\ terms in \lang\ reduce to monostage residuals
% (as opposed to partial values), term splitting at world
% \bbtwo\ (specified by the $\splittwosym$ relation in
% \cref{fig:termSplit}) is simpler than that of \bbonem.

% Consider any $n$-ary \bbtwo-term $e = \scriptCapp {e_1 \ttsemi \ldots
%   \ttsemi e_n}$ (except \texttt{prev}, which is discussed shortly).
% Note that for any $\diatwo e q$, the residual $q$ must have the form
% $\scriptCapp {q_1 \ttsemi \ldots \ttsemi q_n}$, even if $\mathcal{C}$ is
% nullary or the $q_i$ have binders.
% For each $k$, splitting the subterm $e_k$ yields $p_k$ and $l_k.r_k$ where
% \[
% 	\reduce {p_k} {b_k} \text{ and } [{b_k}/{l_k}]{r_k} \equiv {q_k}
% \]
% Correspondingly, splitting $\scriptCapp {e_1 \ttsemi \ldots \ttsemi e_n}$ generates $p$ and $l.r$ where
% \[
% 	\reduce p b \text{ and } [b/l]r \equiv \scriptCapp {q_1 \ttsemi \ldots \ttsemi q_n}
% \]
% Our algorithm accomplishes this by using $p = \tup {p_1, \ldots, p_n}$ and
% $l.r = \tup {l_1, \ldots, l_n}.\scriptCapp {r_1 \ttsemi \ldots \ttsemi r_n}$.

% %Again, this holds even for nullary terms and under binders.
% In particular, $\caseof{e_1}{x_2.e_2}{x_3.e_3}$ splits into:
% \[
% \pipeS {\tup{p_1,p_2,p_3}} {\tup{l_1,l_2,l_3}} {\caseof{r_1}{x_2.r_2}{x_3.r_3}}
% \]
% Note how the precomputations of $e_2$ and $e_3$ are pulled out from
% underneath the binder into a tuple that is always executed in the
% first stage. This occurs regardless of what branch is eventually
% executed in the second stage.

% \paragraph{Staging constructs.}
% Since \texttt{next} and \texttt{prev} do nothing more than convert between 
% second-stage terms at \bbtwo\ and encapsulated second-stage terms at \bbonem,
% their splitting rules are relatively simple.
% Given $\diatwo e q$, splitting yields $p$ and $l.r$ where
% \[
% 	\reduce p b \text{ and } [b/l]r \equiv q
% \]
% Then for $\diaone {\next {e}} {{\hat y} \mapsto q} {\next {\hat y}}$
% and\\ $\rtab {{\hat y} \mapsto q} {\next {\hat y}} \vsplito \mval {\tup{}} {\letin {\hat y} q {\hat y}}$, splitting must produce $c'$ and $l'.r'$ where
% \[
% 	\reduce {c'} {\tup {\tup{},b'}} \text{ and } [b'/l']r' \equiv \letin {\hat y} q {\hat y}
% \]
% Our algorithm chooses $c' = \tup{\tup{},p}$ and $l'.r' = l.r$.
% Applying this process in reverse yields a splitting rule for $\prev e$ which applies \texttt{pi1} to the combined term of $e$.
